### 2018-02-20 submission/manuscript.Rmd
update discussion

Our observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have both significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5]. HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this metric as part of their calculations. Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4]. These differences were observed in high biomass samples, where biases introduced by such components like kit contamination may have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.

Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used. There was few differences within individuals between corresponding 5-cycle increments (e.g. 15 to 20, 20 to 25, etc.). However, there is a clear trend that suggests that 20 cycles is very different then 25 cycles [Figure 3]. We may just not have enough samples to adequately assess the true magnitude difference between 20 and 25 cycle communities and higher cycle numbers analyzed. These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached. For some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on HiFi DNA polymerase. It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases (e.g. Platinum).

Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimera prevalence that cannot be removed using bioinformatic approaches [Figure 4 & 5]. Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014]. In fact, from the material safety data sheet (MSDS), the differences between the two mixes is not immediately apparent. Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pycrococcus* spp GB-D polymerase and a platinum *Taq* antibody. It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate.

>>>>

These observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5]. HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this measure as part of their calculations. Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4]. Importantly, high biomass samples may underestimate the biases present within low biomass samples. An indication that this may be the case is the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples.

Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30 cycles, that are commonly used. There was few differences within individuals between corresponding 5-cycle increments (e.g. 15 to 20, 20 to 25, etc.). However, there is a clear trend that suggests that the community at 20 cycles is different then the community at 25 cycles [Figure 3]. These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached. Further, for some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on HiFi DNA polymerase. It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases.

Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimera prevalence that cannot be removed using bioinformatic approaches [Figure 4 & 5]. Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014]. In fact, from the material safety data sheet (MSDS), the differences between the two is not immediately apparent. Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate.

### 2018-02-20 submission/manuscript.tex
update discussion

Our observations build upon previous studies (Gohl et al., 2016) by
showing that different HiFi DNA polymerases have both significant
differences in the number of OTUs and that changes in total OTUs
correlate with chimeras not removed after sequence processing {[}Figure
1-2 \& 5{]}. HiFi DNA polymerase dependent differences in total number
of OTUs are important to consider since many diversity metrics rely on
this metric as part of their calculations. Our observations show that
HiFi DNA polymerase can have a noticeable affect on the OTUs generated
and these differences are consistent across sub-sampling depth and PCR
cycle number {[}Figure 2-4{]}. These differences were observed in high
biomass samples, where biases introduced by such components like kit
contamination may have less of an effect, suggesting that these
differences may be exacerbated in low biomass samples.

Although we did not observe strong differences in the Bray-Curtis index
the data suggests that there may be differences between 15 and 20 cycles
versus higher cycle numbers, such as 30x, that are commonly used. There
was few differences within individuals between corresponding 5-cycle
increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a clear
trend that suggests that 20 cycles is very different then 25 cycles
{[}Figure 3{]}. We may just not have enough samples to adequately assess
the true magnitude difference between 20 and 25 cycle communities and
higher cycle numbers analyzed. These findings, in conjunction with the
PERMANOVA results, suggest that cycle number can change bacterial
community calculations but that these differences are minimal once 25
cycles are reached. For some HiFI DNA polymerases, increasing the
sub-sampling depth may reduce some of these observed community
differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was strongly correlated with the number of OTUs
which is relied upon heavily for many community metric calculations.
However, Bray-Curtis analysis with PERMANOVA showed few differences
based on HiFi DNA polymerase. It is possible that many of the increased
number of OTUs are not highly abundant allowing the Bray-Curtis index to
be able to successfully down-weight these respective OTUs (Minchin,
1987). The choice of downstream diversity metric could be an important
consideration in helping to mitigate these observed changes due to high
chimera prevalence in specific HiFi DNA polymerases (e.g.~Platinum).

Our observations suggest that there are clear HiFi DNA polymerase
dependent differences in both per base error rate and chimera prevalence
that cannot be removed using bioinformatic approaches {[}Figure 4 \&
5{]}. Although it may be a natural assumption that the variation may be
due to the DNA polymerase family, the highest chimera rate, from
Platinum, was a family A polymerase while the lowest, from Accuprime,
was also an A family polymerase (Ishino \& Ishino, 2014). In fact, from
the material safety data sheet (MSDS), the differences between the two
mixes is not immediately apparent. Both Accuprime and Platinum contain a
recombinant \emph{Taq} DNA polymerase, a \emph{Pycrococcus} spp GB-D
polymerase and a platinum \emph{Taq} antibody. It is possible that
differences in how the recombinant \emph{Taq} was generated could be the
main reason for the differences in chimera rate.

>>>>

These observations build upon previous studies (Gohl et al., 2016) by
showing that different HiFi DNA polymerases have significant differences
in the number of OTUs and that changes in total OTUs correlate with
chimeras not removed after sequence processing {[}Figure 1-2 \& 5{]}.
HiFi DNA polymerase dependent differences in total number of OTUs are
important to consider since many diversity metrics rely on this measure
as part of their calculations. Our observations show that HiFi DNA
polymerase can have a noticeable affect on the OTUs generated and these
differences are consistent across sub-sampling depth and PCR cycle
number {[}Figure 2-4{]}. Importantly, high biomass samples may
underestimate the biases present within low biomass samples. An
indication that this may be the case is the earlier detection of
differences in total number of OTUs between HiFi DNA polymerases in the
mock versus fecal samples.

Although we did not observe strong differences in the Bray-Curtis index
the data suggests that there may be differences between 15 and 20 cycles
versus higher cycle numbers, such as 30 cycles, that are commonly used.
There was few differences within individuals between corresponding
5-cycle increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a
clear trend that suggests that the community at 20 cycles is different
then the community at 25 cycles {[}Figure 3{]}. These findings, in
conjunction with the PERMANOVA results, suggest that cycle number can
change bacterial community calculations but that these differences are
minimal once 25 cycles are reached. Further, for some HiFI DNA
polymerases, increasing the sub-sampling depth may reduce some of these
observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was strongly correlated with the number of OTUs
which is relied upon heavily for many community metric calculations.
However, Bray-Curtis analysis with PERMANOVA showed few differences
based on HiFi DNA polymerase. It is possible that many of the increased
number of OTUs are not highly abundant allowing the Bray-Curtis index to
be able to successfully down-weight these respective OTUs (Minchin,
1987). The choice of downstream diversity metric could be an important
consideration in helping to mitigate these observed changes due to high
chimera prevalence in specific HiFi DNA polymerases.

Our observations suggest that there are clear HiFi DNA polymerase
dependent differences in both per base error rate and chimera prevalence
that cannot be removed using bioinformatic approaches {[}Figure 4 \&
5{]}. Although it may be a natural assumption that the variation may be
due to the DNA polymerase family, the highest chimera rate, from
Platinum, was a family A polymerase while the lowest, from Accuprime,
was also an A family polymerase (Ishino \& Ishino, 2014). In fact, from
the material safety data sheet (MSDS), the differences between the two
is not immediately apparent. Both Accuprime and Platinum contain a
recombinant \emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D
polymerase and a platinum \emph{Taq} antibody. It is possible that
differences in how the recombinant \emph{Taq} was generated could be the
main reason for the differences in chimera rate.
